# Describing the Observations

## Describing nominal data

Nominal data are produced by assigning the subjects to one of a set of categories, such as dead/alive, germinated/ungerminated, red/blue/green, and so on. For this data, especially when the number of categories is higher than two, the statistics described in book Chapter 3 cannot be used and different approaches need to be sought for their description.

The most widespread technique is to derive **absolute  frequencies**, that are the counts of individuals in each category; by doing so, a  **distribution of frequencies** is built. 

As an example of nominal data we can take the 'mtcars' dataset, that was extracted from the 1974 Motor Trend US magazine and comprises 32 old automobiles. The dataset is available in R and we show part of it in  table 3.2.


```{r echo=FALSE}
data(mtcars)
knitr::kable(mtcars[,c(8,10)], caption = "Part of the dataset 'mtcars' in R, representing the characteristics of 32 old automobiles; 'vs' is the type of engine (0 for V-shaped and 1 for straight engine) and 'gear' is the number of forward gears. More detail is given in the text.", booktabs = T)
```

The variable 'vs' in 'mtcars' takes the values 0 for V-shaped engine and 1 for straight engine. Obviously, the two values 0 and 1 are just used to name the two categories and the resulting variable is purely nominal. The absolute frequencies of cars in the two categories are, respectively 18 and 14 and they are easily obtained by a counting process. In R, these absolute frequencies can be obtained by the `table()` function, as we show in the box below.


```{r}
data(mtcars)
table(mtcars$vs)
```

The **relative frequencies** are obtained by dividing the absolute frequencies by the total number of observations. These frequencies are, respectively, 0.5625 and 0.4375 and, in R, they can be obtained by the joint usage of the `table()` and `length()` functions, where this latter returns the number of values in a vector.

```{r}
table(mtcars$vs)/length(mtcars$vs)
```

If we consider a variable where the classes can be logically ordered, we can also calculate the **cumulative frequencies**, by summing up the frequency for one class with the frequencies for all previous classes. As an example we take the 'gear' variable in the 'mtcars' dataset, showing the number of forward gears for each car. We can easily see that 15 cars have 3 gears and 27 cars have 4 gears or less. In R, cumulative sums can be calculated by using the `cumsum()` function, as shown in the box below.


```{r}
cumsum(table(mtcars$gear))
```

## Binning/Bucketing

In some circumstances, it may be convenient to 'bin' a continuous variable into a set of intervals. For example, if we have recorded the ages of a big group of people, we can divide the scale into intervals of five years (e.g., from 10 to 15, from 15 to 20 and so on) and, eventually, assign each individual to the appropriate age class. Such a technique is called **binning** or **bucketing**.

As an example, we can consider the 'co2' dataset, that is included in the base R installation. It contains 468 values of CO_2_ atmospheric concentrations, expressed in parts per million, as observed at monthly intervals in the US. With such a big dataset, the mean and standard deviation are not sufficient to get a good feel for the data and it would be important to have an idea of the shape of the dataset. Therefore we can split the continuous scale into a series of intervals, from 310 ppm to 370 ppm, with breaks every 10 ppm and count the observations in each interval. In the box below, the function `cut()` assigns each value to the corresponding interval (please note the 'breaks' argument, which sets the margins of each interval. Intervals are, by default, left open and right-closed), while the function `table()` calculates the frequencies.


```{r}
data(co2)
co2 <- as.vector(co2)
mean(co2)
min(co2)
max(co2)

# binning
classes <- cut(co2, breaks = c(310,320,330,340,350,360,370))
freq <- table(classes)
freq
```


## Descriptive stats for distributions of frequencies

For nominal data, we can retrieve the **mode**, which is the class with the highest frequency. For ordinal data, wherever distances between classes are meaningful, and for discrete data, we can also calculate the median and other percentiles, as well as the mean and other statistics of spread (e.g., variance, standard deviation). The mean is calculated as:

$$\mu = \frac{\sum\limits_{i = 1}^n f_i x_i}{\sum\limits_{i = 1}^n f_i}$$

where $x_i$ is the value for the i-th class, and $f_i$ is the frequency for the same class. Likewise, the deviance, is calculated as:

$$SS = \sum\limits_{i = 1}^n f_i (x_i - \mu)^2$$

For example, considering the 'gear' variable in Table 3.2, the average number of forward gears is:

$$\frac{ 15 \times 3 + 12 \times 4 + 5 \times 5}{15 + 12 + 5} = 3.6875$$

while the deviance is:

$$SS = 15 \times (3 - 3.6875)^2 + 12 \times (4 - 3.6875)^2 + 5 \times (5 - 3.l875)^2 = 16.875$$

With interval data (binned data), descriptive statistics should be calculated by using the raw data, if they are available. If they are not,  we can use the frequency distribution obtained from binning, by assigning to each individual the central value of the interval class to which it belongs. As an example, we can consider the distribution of frequencies in Table 3.3, relating to the time (in minutes) taken to complete a statistic assignment for a group of students in biotechnology. We can see that the mean is equal to:

$$\frac{7.5 \times 1 + 12.5 \times 4 + 17.5 \times 3 + 22.5 \times 2}{10} = 15.5$$

```{r echo=FALSE}
time <- c("5 - 10", "10 - 15", "15 - 20", "20 - 25")
center <- c(7.5, 12.5, 17.5, 22.5)
count <- c(1, 4, 3, 2)
tab <- data.frame("Time interval" = time, "Central value" = center,
                  "Count" = count, check.names = F)
knitr::kable(tab, caption = "Distribution of frequency for the time (in minutes) taken to complete a statistic assignment for a group of students in biotechnology", align = "ccc", booktabs = T)
```


The calculation of the deviance is left as an exercise.

## Graphical representations of nominal data

A distribution of frequency can be represented by using a barplot, as described in the main book. It can also be represented by using a 'pie' chart, drawn with `ggplot`. The coding consists of producing a bar chart with stacked segments and rotating the coordinate system, as shown in the box below.


```{r}
library(ggplot2)
dfr <- data.frame("Class" = names(freq),
                  "Freq" = as.numeric(freq))
ggplot(data = dfr) +
  geom_bar(aes(x = NA, y = Freq, fill = Class),
           stat="identity", position = "fill") +
  coord_polar("y", start=0) +
  theme_void() # remove background, grid, numeric labels
```


## Contingency tables

When we have more than one cataegorical variable, we can summarise the distribution of frequency by using two-way tables, usually known as **contingency tables**  or crosstabs. These tables can be created by providing two or more vectors as arguments to the `table()` function; it is important to keep in mind that, even if the resulting table may resamble a matrix or a data.frame, a contingency table represents a peculiar class in R, with several specific methods, that we will explore later. The box below shows a contingency table that shows the cross tabulation for 'gear' and 'vs' in 'mtcars'.

```{r}
table(mtcars$vs, mtcars$gear)
```

Another interesting table is contained in the 'datasets' package and is named  'HairEyeColor'. It shows the distribution of hair and eye color in 592 statistics students, depending on sex; both characters are expressed in four classes, i.e. black, brown, red and blond hair and brown, blue, hazel and green eyes. Considering females, the contingency table is reported in the following table and it is augmented with row and column sums (see later).


```{r echo=FALSE}
data(HairEyeColor)
tab <- HairEyeColor[,,2]
tab <- matrix(as.numeric(tab), 4, 4)
# tab <- cbind(tab, apply(tab, 1, sum))
# tab <- rbind(tab, apply(tab, 2, sum))

rn <- c("Black hair", "Brown hair", "Red hair", "Blond hair")
cn <- c("Brown eye", "Blue eye", "Hazel eye", "Green eye")
row.names(tab) <- rn
colnames(tab) <- cn
knitr::kable(tab, align = "cccc", caption = "Distribution of hair and eye color for 313 female statistics students, augmented with row and column sums. Dataset taken from R package 'datasets'", booktabs = T)
```


### Independence

With a contingency table, we may be interested in assessing whether the two variables show some sort of dependency relationship. In the previous example, is there any relationship between the color of the eyes and the color of the hair? If not, we say that the two variables are independent. Independency is assessed by using the $\chi^2$ statistic.

In order to make hand calculations easier, let's restrict our attention to people with Black/Blond hair and Brown/Blue eyes. This 'restricted' contingency table is reported below and clearly shows that the vast majority of people with black hair have brown eyes, while the vast majority of people with blond hair have blue eyes. It would apper that there is a strong association between the color of hair and the color of eyes.

```{r echo=FALSE}
data(HairEyeColor)
tab <- HairEyeColor[c(1,4),c(1,2),2]
tab <- matrix(as.numeric(tab), 2, 2)
tab <- cbind(tab, apply(tab, 1, sum))
tab <- rbind(tab, apply(tab, 2, sum))

rn <- c("Black hair", "Blond hair", "COLUMN SUMS")
cn <- c("Brown eye", "Blue eye", "ROW SUMS")
row.names(tab) <- rn
colnames(tab) <- cn
knitr::kable(tab, align = "ccc", caption = "Excerpt from the previous table, considering (for the sake of simplicity) only two color classes for both hair and eyes", booktabs = T)
```

In order to elucidate such an association, we can, as the first step calculate the *marginal frequencies*, i.e. the sums of frequencies by row and by column (please note that the entries of a contingency table are called *joint frequencies*). These sums are reported in the Table above.

```{r eval=FALSE, include=FALSE}
# rowSum <- apply(tab, 1, sum)
# colSum <- apply(tab, 2, sum)
# 
# tabSum <- rbind(cbind(tab, colSum), c(rowSum, sum(tab)))
# library(knitr)
# kable(tabSum)
# tabSum[1,] * 122/313
# exp <- read.table(pipe("pbpaste"), header=T)
# kable(exp)
# sum( ((tab - exp)/exp)^2 )
# summary(tab)
```

In total, we have 45 individuals with black hair, that is 45/113 × 100 = 39.8% of the whole sample, while we have 68/113×100 = 60.2% of individuals with blond hair. If the two characters were really independent, the above percentages should not change, depending on the color of eyes; consequently, in the brown eyes group, the number of individuals with black hair should be 15.93 (i.e., 39.8% of 40), while the number of individuals with blonde hair should be 24.08 (i.e. 60.2% of 40). From there, we could derive a full table of expected frequencies, assuming complete independence between the two characters (see below).

```{r echo=FALSE}
tab <- matrix(c(15.93, 29.07, 24.07, 43.93), 2, 2, byrow = TRUE)
row.names(tab) <- c("Black", "Blonde")
colnames(tab) <- c("Brown", "Blue")
knitr::kable(tab, align = "ccccc", caption = "Expected counts of hair and eye color, assuming total lack of dependency between the two characters.", booktabs = T)
```


The observed and expected values are different, which might indicate a some sort of relationship between the two variables; for example, having red hair might imply that we are more likely to have eyes of a certain color. In order to quantify the discrepancy between the two tables, we calculate the $\chi^2$ stat, that is:

$$\chi ^2  = \sum \left[ \frac{\left( {f_o  - f_e } \right)^2 }{f_e } \right]$$

where $f_o$ are the observed frequencies and $f_e$ are the expected frequencies. For this example:

$$\chi ^2  = \left[ \frac{\left( {36  - 15.93 } \right)^2 }{15.93} \right] + \left[ \frac{\left( {9  - 29.07 } \right)^2 }{129.07} \right] +\left[ \frac{\left( {4  - 24.07 } \right)^2 }{24.07 } \right] + \left[ \frac{\left( {64  - 43.93 } \right)^2 }{43.93 } \right] = 65.05$$


The final $\chi^2$ value should be equal to 0 in case of independence and it should increase as the relationship between the two variables increases. With R, the chi square value is provided as an output of the `summary()` method, applied to the table object.

```{r}
tab <- as.table(matrix(c(36, 9, 4, 64), 2, 2, byrow = TRUE))
summary(tab)
```

The $\chi^2$ value needs to be compared with its maximum allawable value, which is:

$$\max \left( \chi ^2 \right)  = n \cdot \min (r - 1,\,c - 1)$$

i.e. the product between the number of subjects ($n$) and the minimum value between the number of rows minus one and the number of columns minus one (in our case, it is $113 \times 1 = 113$). This suggests that the two variables are not independent. The square root of the ratio between the observed chi square value and the maximum value is know as the Cramer's V coefficient and it is more understandable than the chi square, as it is always included between 0 and 1.

```{r}
Vcram <- sqrt(summary(tab)$statistic/(sum(tab) * min(nrow(tab) - 1, ncol(tab) - 1)))
Vcram
```


## Questions and exercises

1. List and discuss the main statistics for nominal and ordinal variables
2. Describe and discuss briefly the chi square coefficient
3. A scientist ha compared the proportion of females in two random samples of insects treated with two different substances (A and B). With substance A, they found 275 males and 175 females, while, with substance B they found 326 males and 297 females. Considering these results, determine the degree of association between the two variables (sex and substance), in relation to the maximum and minimum allowable level for the selected statistic.
4. Load the csv file 'students.csv' from [https://www.casaonofri.it/_datasets/students.csv](https://www.casaonofri.it/_datasets/students.csv). This dataset relates to a number of students, their votes in several undergraduate exams and information on high school. Determine: (i) the absolute and relative frequencies for the different subjects; (ii) the frequency distribution of votes in three classes (bins): <24, 24-27, >27; (iii) whether the votes depend on the exam subject and (iv) whether the votes depend on the high school type.

